---
title: "Tropical fish stock assessment with R"
author: "Tobias Mildenberger, Marc Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TropFishR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes the functions of the package "TropFishR" in detail, puts them into context and explains strenghts and shortcomings. 
The package TropFishR is a compilation of fish stock assessment methods and models listed in the FAO Technical Manual "Introduction into tropical fish stock assessment" by Sparre and Venema (1998) and its exercise book (Spaare & Venema 1999).



## Contents

1. Growth parameters

    * `Bhattacharya.R`
    
    * `ELEFAN.R`

1. Mortality parameters

    * `Z_CPUE.R`
   
    * `CatchCurve.R`
   
    * `BevertonHoltsZ.R`
   
    * `PaulysM.R`
   
    * `RikhterEfanovsM.R`
   
1. Virtual population dynamics

    * `VPA.R`

1. Gear selectivity

    * `select.R`
    
    * `select_ogive.R`
    
    * `rtypes_Millar.R`
    
    * `select_Millar.R`

1. Prediction models

    * `Predict_BevertonHolt.R`

    * `Predict_ThompsonBell1.R`
    
    * `Predict_ThompsonBell2.R`
    
    * `Predict_ThompsonBell_Fleets.R`
    
    * `Predict_ThompsonBell_FM_Lc.R`

1. Surplus production models

    * `SchaeferFox.R`




## 1. Growth parameters

### Bhattacharya method

### ELEFAN


## 2. Mortality parameters



## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
