
---
title: "Single-species fish stock assessment with TropFishR"
author: "Tobias K. Mildenberger"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{tutorial_stock_assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: TropFishR.bib
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

This tutorial illustrates the application of the **`TropFishR`** package to perform a single-species fish stock assessment with length frequency (LFQ) data. According to @Sparre1998b, this includes following steps: (1) estimation of biological stock characteristics (growth and natural mortality), (2) exploration of fisheries aspects (exploitation rate and selectivity), (3) assessment of stock size and status. The order of the methods is important as they build upon each other in a sequential way. Data from literature might be used to skip a step in this workflow or to compare them to the outcomes of this routine.

### *Installing TropFishR*

The current version of **`TropFishR`** (v1.6.3) requires R $>= 3.0.0$ and can be downloaded from CRAN as follows.

```{r, echo = TRUE, include = TRUE, eval = FALSE}
install.packages("TropFishR", repos = "https://cran.rstudio.com/")
```

Alternatively, the package can be installed from GitHub:

```{r, eval=FALSE, echo=TRUE}
install.packages("remotes")
remotes::install_github("tokami/TropFishR")
```

The package is loaded into the R environment with:

```{r, eval=TRUE, echo=TRUE}
library(TropFishR)
```

The tutorial will make use of a synthetic LFQ data set included in the package
("synLFQ7"). We load the data set into the R environment with `{r eval=FALSE}
data("synLFQ7")` and rename it with: `{r, eval=FALSE} lfq <- synLFQ7`.

```{r, eval=TRUE, echo=FALSE}
data("synLFQ7")
lfq <- synLFQ7
```



### *Biological stock characteristics*

Growth, natural mortality, recruitment patterns and the stock-recruitment relationship are important biological stock characteristics and input parameters for population dynamics and yield per recruit models.

#### *Growth parameters*

Commonly used growth parameters are the asymptotic length ($L_{inf}$), the growth coefficient ($K$) and the theoretical length at age zero ($t_{0}$) of the von Bertalanffy growth function (VBGF). The ELEFAN (ELectronic LEngth Frequency ANalysis) methods allow to estimate $L_{inf}$ and $K$ from LFQ data by restructuring the data and fitting growth curves through the restructured LFQ data [@Pauly1980].
I recommend to start by visualising the raw and restructured LFQ data, which aids in determining an appropriate bin size and the moving average of the restructuring procedure. The function "lfqModify" allows to change the bin size by setting the argument "bin_size" to a numeric. The function "lfqRestructure" is used for the restructuring process, where the argument "MA" allows to control the number of bins used for the moving average and the argument "addl.sqrt" allows to apply an additional squareroot transformation in the restructuring process, which reduces the weighting of large individuals.

```{r Figure 1, echo=TRUE, eval=TRUE, fig.cap="Length frequency data visualised in terms of (a) catches and (b) restructured data with MA = 7."}
## set seed value for reproducible results
set.seed(1)

## adjust bin size
lfq_bin2 <- lfqModify(lfq, bin_size = 2)

## plot raw and restructured LFQ data
ma <- 7
lfq_bin2_res <- lfqRestructure(lfq_bin2, MA = 7, addl.sqrt = FALSE)

opar <- par(mfrow = c(2,1), mar = c(2,5,2,3), oma = c(2,0,0,0))
plot(lfq_bin2_res, Fname = "catch", date.axis = "modern")
plot(lfq_bin2_res, Fname = "rcounts", date.axis = "modern")
par(opar)
```

Based on visual inspection of the length-frequency distributions, a bin size of
2 cm and a moving average of 7 seems appropriate for this data set and will be
used in the following.

In **`TropFishR`**, there are 4 different methods based on the ELEFAN
functionality: (i) K-Scan for the estimation of K for a fixed value of
$L_{inf}$, (ii) Response Surface Analysis (RSA), (iii) ELEFAN with simulated
annealing ("ELEFAN_SA"), and (iv) ELEFAN with a genetic algorithm ("ELEFAN_GA"),
where the last three methods all allow to estimate K and $L_{inf}$
simultaneously. The K-Scan method does not allow to test if different
combinations of $L_{inf}$ and K might result in a better fit. RSA with a range
around $L_{inf}$ can be used to check different combinations. However, RSA can
be very slow and does not allow to optimise over the parameters C and $t_s$ of
the seasonalised VBGF (soVBGF). It only allows to compare the score of ELEFAN
runs with manually fixed C and $t_s$ values. In contrast, the newly implemented
ELEFAN method `ELEFAN_SA` using a simulated annealing algorithm [@Xiang2013] and
`ELEFAN_GA` using genetic algorithms allow for the optimisation of the soVBGF
[@Taylor2016]. The optimisation procedure in the simulated annealing algorithm
gradually reduces the stochasticity of the search process as a function of the
decreasing "temperature" value, which describes the probability of accepting
worse conditions.

Here, we will apply the two new ELEFAN functions `ELEFAN_SA` and `ELEFAN_GA`.
Both functions require to define the lower (`low_par`) and upper (`up_par`)
bounds of the search space for all growth parameters. For this data set, we
assume a wide search space for $K$ and use the maximum length in the data to
define the search space for $L_{\infty}$ [@Taylor1958; @Beverton1963]. We assume
the full parameter space for the other parameters ($t_{anchor}$, $C$, $t_s$).

```{r elefan_search_space, echo=TRUE, eval=TRUE}
## coarse estimate of Linf
linf_guess <- max(lfq_bin2$midLengths) / 0.95

## lower search space bounds
low_par <- list(Linf = 0.8 * linf_guess,
                K = 0.01,
                t_anchor = 0,
                C = 0,
                ts = 0)

## upper search space bounds
up_par <- list(Linf = 1.2 * linf_guess,
               K = 1,
               t_anchor = 1,
               C = 1,
               ts = 1)
```

```{r Figure 4, fig.height=5, fig.width=5, echo=TRUE, eval=TRUE, results="hide", fig.cap="Score graph of the ELEFAN method with simulated annealing. Green dots indicate the runnning minimum value of the cost function, while blue dots indicate the mean score of each iteration. The red line shows the decline of the 'temperature' value, which describes the probability of accepting worse solutions as the parameter space is explored."}
## run ELEFAN with simulated annealing
res_SA <- ELEFAN_SA(lfq_bin2, SA_time = 60*0.5, SA_temp = 6e5,
                   MA = ma, seasonalised = TRUE, addl.sqrt = FALSE,
                   init_par = list(Linf = linf_guess,
                                   K = 0.5,
                                   t_anchor = 0.5,
                                   C=0.5,
                                   ts = 0.5),
                   low_par = low_par,
                   up_par = up_par)

## show results
res_SA$par
res_SA$Rn_max
```


Note that the computing time can be controlled with the argument "SA_time" and
the results might change when increasing the time, in case the stable optimum of
the objective function was not yet reached^[Stable optimum is indicated by
                                            overlapping blue and green dots in the score graph.].
Due to the limitations of
the vignette format the computation time was set to 0.5 minutes, which results
already in acceptable results of $L_{inf}$ = `r {round(res_SA$par$Linf,2)}`, K =
  `r {round(res_SA$par$K,2)}`, $t_{anchor}$ = `r {round(res_SA$par$t_anchor,2)}`,
C = `r {round(res_SA$par$C,2)}`, and $t_s$ = `r {round(res_SA$par$ts,2)}` with a
score value ($Rn_{max}$) of `r {round(res_SA$Rn_max,2)}`. I recommend to
increase 'SA_time' to 3 - 5 minutes to increase chances of finding the stable
optimum. Another new optimisation routine is based on generic algorithms and is
applied by:
  


```{r Figure 5, fig.height=5, fig.width=5, eval=TRUE, results="hide", fig.cap="Score graph of the ELEFAN method with genetic algorithms. Green dots indicate the runnning maximum value of the fitness function, while blue dots indicate the mean score of each iteration."}
## run ELEFAN with genetic algorithm
res_GA <- ELEFAN_GA(lfq_bin2, MA = ma, seasonalised = TRUE,
                    maxiter = 50, addl.sqrt = FALSE,
                    low_par = low_par,
                    up_par = up_par,
                    monitor = FALSE)

## show results
res_GA$par
res_GA$Rn_max
```

The generation number of the `ELEFAN_GA` was set to only 50 generations
(argument 'maxiter'), which returns following results: $L_{inf}$ = `r
{round(res_GA$par$Linf,2)}`, K = `r {round(res_GA$par$K,2)}`, $t_{anchor}$ = `r
{round(res_GA$par$t_anchor,2)}`, C = `r {round(res_GA$par$C,2)}`, and $t_s$ = `r
{round(res_GA$par$ts,2)}` with a score value ($Rn_{max}$) of `r
{round(res_GA$Rn_max,2)}`. As with `ELEFAN_SA` the generation number was hold
down due to the vignette format and should be increased in order to find more
stable results. According to [@Pauly1980] it is not possible to estimate $t_{0}$
(theoretical age at length zero) from LFQ data alone. However, this parameter
does not influence results of the methods of the traditional stock assessment
workflow (catch curve and yield per recruit model) and can be set to
zero (Mildenberger, unpublished). The ELEFAN methods in this package do not
return starting points as FiSAT II users might be used to. Instead, they return
the parameter "t_anchor", which describes the fraction of the year where yearly
repeating growth curves cross length equal to zero; for example a value of 0.25
refers to April 1st of any year. The maximum age is estimated within the ELEFAN
function: it is the age when length is 0.95 $L_{inf}$. However, this value can
also be fixed with the argument "agemax", when alternative information about the
maximum age of the fish species is available.

The jack knife technique allows to estimate a confidence interval around the
parameters of the soVBGF [@Quenouille1956; @Tukey1958; @Tukey1986]. This can be
automated in R with following code:

```{r, eval = FALSE, echo = TRUE}
## list for results
JK <- vector("list", length(lfq_bin2$dates))

## loop
for(i in 1:length(lfq_bin2$dates)){
  loop_data <- list(dates = lfq_bin2$dates[-i],
                  midLengths = lfq_bin2$midLengths,
                  catch = lfq_bin2$catch[,-i])
  tmp <- ELEFAN_GA(loop_data, MA = ma, seasonalised = TRUE,
                    maxiter = 50, addl.sqrt = FALSE,
                    low_par = low_par,
                    up_par = up_par,
                    monitor = FALSE, plot = FALSE)
  JK[[i]] <- unlist(c(tmp$par, list(Rn_max=tmp$Rn_max)))
}

## bind list into dataframe
JKres <- do.call(cbind, JK)

## mean
JKmeans <- apply(as.matrix(JKres), MARGIN = 1, FUN = mean)

## confidence intervals
JKconf <- apply(as.matrix(JKres), MARGIN = 1, FUN = function(x) quantile(x, probs=c(0.025,0.975)))
JKconf <- t(JKconf)
colnames(JKconf) <- c("lower","upper")

## show results
JKconf
```

Depending on the number of sampling times (columns in the catch matrix) and the
"maxiter",, this loop can take some time as ELEFAN runs several times, each time
removing the catch vector of one of the sampling times.

The fit of estimated growth parameters can also be explored visually and
indicates high similarity with true growth curves and a good fit through the
peaks of the LFQ data.


```{r Figure 6, echo = TRUE, fig.cap="Graphical fit of estimated and true growth curves plotted through the length frequency data. The growth curves with the true values are displayed in grey, while the blue and green curves represent the curves of ELEFAN_SA and ELEFAN_GA, respectively."}
## plot LFQ and growth curves
plot(lfq_bin2_res, Fname = "rcounts",date.axis = "modern", ylim=c(0,130))
lt <- lfqFitCurves(lfq_bin2, par = list(Linf=123, K=0.2, t_anchor=0.25, C=0.3, ts=0),
                   draw = TRUE, col = "grey", lty = 1, lwd=1.5)
lt <- lfqFitCurves(lfq_bin2, par = res_SA$par,
                  draw = TRUE, col = "darkblue", lty = 1, lwd=1.5)
lt <- lfqFitCurves(lfq_bin2, par = res_GA$par,
                   draw = TRUE, col = "darkgreen", lty = 1, lwd=1.5)
```

For further analysis, we use the outcomes of ELFAN with genetic algorithm by adding them to the Thumbprint Emperor data list.

```{r}
## assign estimates to the data list
lfq_bin2 <- lfqModify(lfq_bin2, par = res_GA$par)
```



#### *Natural mortality rate*

The instantaneous natural mortality rate (M) is an influential parameter of
stock assessment models and its estimation is challenging [@Kenchington2014;
@Powers2014]. When no controlled experiments or tagging data is available the
main approach for its estimation is to use empirical formulas. Overall, there
are at least 30 different empirical formulas for the estimation of this
parameter [@Kenchington2014] relying on correlations with life history
parameters and/or environmental information. We apply the most recent formula,
which is based upon a meta-analysis of 201 fish species [@Then2015]. This method
requires estimates of the VBGF growth parameters [$L_{inf}$ and K; @Then2015].


```{r, echo=TRUE}
## estimation of M
Ms <- M_empirical(Linf = lfq_bin2$par$Linf, K_l = lfq_bin2$par$K, method = "Then_growth")
lfq_bin2$par$M <- as.numeric(Ms)

## show results
paste("M =", as.numeric(Ms))
```

The result is a natural mortality of `r {round(as.numeric(Ms),2)}` year$^{-1}$.


### *Total and fishing mortality rates*

In order to estimate the level of exploitation, knowledge on fishing mortality (F) (usually derived by subtracting natural mortality from total mortality) and gear selectivity is necessary. The length-converted catch curve allows the estimation of the instantaneous total mortality rate (Z) of LFQ data and the derivation of a selection ogive. Here we skip an in-depth selectivity exploration, because more data would be required for this assessment^[For a comprehensive description of selectivity estimation refer to @Millar1997b.]. The following approach assumes a logistic selection ogive, typical for trawl-net selectivity, which may provide an appropriate first estimate in the case of LFQ data derived from a mixture of gears.
Total mortality rate is estimated with a sample of the catch representative for
the whole year. Besides, changing the bin size, the function `lfqModify` allows
to rearrange the catch matrix in the required format (catch vector per year) and
to pool the largest length classes with only a few individuals into a plus group
(necessary later for the cohort analysis). As with the Powell-Wetherall method,
the `reg_int` argument is necessary to avoid the interactive plotting function
(more information in `help(catchCurve)`). The argument `calc_ogive` allows the
estimation of the selection ogive.

```{r}

## define plus group as largest length class smaller than Linf
plus_group <- lfq_bin2$midLengths[max(which(lfq_bin2$midLengths < lfq_bin2$par$Linf))]

## summarise catch matrix into vector and add plus group
lfq_catch_vec <- lfqModify(lfq_bin2, vectorise_catch = TRUE, plus_group = plus_group)

```


CHECK:
The argument `catch_columns` allows
to choose the columns of the catch matrix which will be summarised for the
analysis. Here all columns are used as the catch matrix only includes catches
from 2016. If data of several years are available, the data can be aggregated
yearly and the results can be averaged or the data of several years is analysed
jointly assuming constant growth parameters.

The argument `reg_int` is necessary in this tutorial because the
"powell_wetherall" function includes an interactive plotting function where
points for the regression analysis have to be selected by the user. Typically,
one would not use this argument and instead choose which points to include in
the regression analysis by clicking on the interactive plot (for more
information see `help(powell_wetherall)`).


```{r Figure 7,echo=TRUE, fig.width=6, fig.height=5, fig.cap="Catch curve with selected points for the regression analysis and in the second panel the selection ogive with age at first capture.", message = FALSE, warning=FALSE}

## run catch curve
res_cc <- catchCurve(lfq_catch_vec, reg_int = c(18,55), calc_ogive = TRUE)

## assign estimates to the data list
lfq_catch_vec$par$Z <- res_cc$Z
lfq_catch_vec$par$FM <- as.numeric(lfq_catch_vec$par$Z - lfq_catch_vec$par$M)

```

```{r, echo=FALSE, eval=TRUE}
paste("Z =",round(lfq_catch_vec$par$Z,2))
paste("FM =",round(lfq_catch_vec$par$FM,2))
paste("L50 =",round(res_cc$L50,2))
```

The catch curve analysis returns a Z value of `r {round(lfq_catch_vec$par$Z,2)}` $year^{-1}$. By subtracting M from Z, the fishing mortality rate is derived: `r {round(lfq_catch_vec$par$FM,2)}` $year^{-1}$. The selectivity function of the catch curve estimated a length at first capture ($L_{50}$) of `r {round(res_cc$L50,2)}` cm.

### *Stock status*

#### *Exploitation rate*

The exploitation rate is defined as $E = F/Z$ and in this example  Gulland (1950).

```{r, echo=FALSE, eval=TRUE}
lfq_catch_vec$par$E <- lfq_catch_vec$par$FM / lfq_catch_vec$par$Z
paste("E =",round(lfq_catch_vec$par$E,2))
```



#### *Yield per recruit modelling*

Prediction models (or per-recruit models, e.g. Thompson and Bell model) allow to
evaluate the status of a fish stock in relation to reference levels and to infer
input control measures, such as restricting fishing effort or regulating gear
types and mesh sizes. By default the Thompson and Bell model assumes knife edge
selection ($L_{25}$ = $L_{50}$ = $L_{75}$)^[Note that the length at capture has
2 abbreviations $L_{50}$ and $L_c$.]; however, the parameter `s_list` allows for
changes of the selectivity assumptions. The parameter `FM_change` determines the
range of the fishing mortality for which to estimate the yield and biomass
trajectories. In the second application of this model, the impact of mesh size
restrictions on yield is explored by changing $L_{c}$ (`Lc_change`) and F
(`FM_change`, or exploitation rate, `E_change`) simultaneously. The resulting
estimates are presented as an isopleth graph showing yield per recruit. By
setting the argument `stock_size_1` to 1, all results are per recruit. If the
number of recruits (recruitment to the fishery) are known, the exact yield and
biomass can be estimated. The arguments `curr.E` and `curr.Lc` allow to derive
and visualise yield and biomass (per recruit) values for current fishing
patterns.

```{r}
## assign length-weight parameters to the data list
lfq_catch_vec$par$a <- 0.015
lfq_catch_vec$par$b <- 3

## assign maturity parameters
lfq_catch_vec$par$Lmat <- 35
lfq_catch_vec$par$wmat <- 8


## list with selectivity parameters
selectivity_list <- list(selecType = "trawl_ogive",
                         L50 = res_cc$L50, L75 = res_cc$L75)
```

```{r Figure 9, echo=TRUE, eval=TRUE, fig.cap="Results of the Thompson and Bell model: (a) Curves of yield and biomass per recruit. The black dot represents yield and biomass under current fishing pressure. The yellow and red dashed lines represent fishing mortality for maximum sustainable yield (Fmax) and fishing mortality to fish the stock at 50% of the virgin biomass (F0.5). (b) exploration of impact of different exploitation rates and Lc values on the relative yield per recruit."}
## Thompson and Bell model with changes in F
TB1 <- predict_mod(lfq_catch_vec, type = "ThompBell",
                   FM_change = seq(0,1.5,0.05),
                   stock_size_1 = 1,
                   curr.E = lfq_catch_vec$par$E,
                   s_list = selectivity_list,
                   plot = FALSE, hide.progressbar = TRUE)

## Thompson and Bell model with changes in F and Lc
TB2 <- predict_mod(lfq_catch_vec, type = "ThompBell",
                   FM_change = seq(0,1.5,0.1),
                   Lc_change = seq(25,50,0.1),
                   stock_size_1 = 1,
                   curr.E = lfq_catch_vec$par$E,
                   curr.Lc = res_cc$L50,
                   s_list = selectivity_list,
                   plot = FALSE, hide.progressbar = TRUE)

## plot results
par(mfrow = c(2,1), mar = c(4,5,2,4.5), oma = c(1,0,0,0))
plot(TB1, mark = TRUE)
mtext("(a)", side = 3, at = -0.1, line = 0.6)
plot(TB2, type = "Isopleth", xaxis1 = "FM", mark = TRUE, contour = 6)
mtext("(b)", side = 3, at = -0.1, line = 0.6)

## Biological reference levels
TB1$df_Es

## Current yield and biomass levels
TB1$currents
```

Please note that the resolution of the $L_c$ and F changes is quite low and the
range quite narrow due to the limitations in computation time of the vignette
format. The results indicate that the fishing mortality of this example (F = `r
{round(lfq_catch_vec$par$FM,2)}`) is higher than the maximum fishing mortality
($F_{max} =$ `r {round(TB1$df_Es$Fmax,2)}`), which confirms the indication of
the slightly increased exploitation rate (E = `r
{round(lfq_catch_vec$par$E,2)}`). The prediction plot shows that the yield could
be increased when fishing mortality and mesh size is increased. The units are
grams per recruit.


## *Summary*

For management purposes, fish stock assessments are mainly conducted for single species or stocks, which describe the manamgent units of a population. There is much to be gained from multi-species and ecosystem models, but data requirements and complexity make them often unsuitable for deriving management advice. For data-poor fisheries, a traditional fish stock assessment solely based on length-frequency (LFQ) data of one year (as presented here) is particularly useful. LFQ data comes with many advantages over long time series of catch and effort or catch-at-age data [@Mildenberger2016].
In this exercise, the exploitation rate and results of the yield per recruit models indicate that the fiseries is close to sustainable exploitation. The exploration of stock status and fisheries characteristics can of course be extended, but go beyond the scope of this tutorial, which is thought to help getting started with the **`TropFishR`** package. Further details about functions and their arguments can be found in the help files of the functions (`help(...)` or `?..`, where the dots refer to any function of the package). Also the two publications by @Mildenberger2016 and by @Taylor2016 provide more details about the functionality and context of the package.


## *Author's comment*

If you have comments or questions please write an [email](mailto:t.k.mildenberger@gmail.com) or post an issue at [GitHub](https://github.com/tokami/TropFishR/issues). You can follow the development of **`TropFishR`** on [ResearchGate](https://www.researchgate.net/project/TropFishR?_esc=homefeed&_viewIds%5B0%5D=hVPmo2RDCok60qGq2z0JOroN&_viewIds%5B1%5D=zyj0j6jnUgNvrPXMrG9rouAP).


## *References*
